---
layout: page
title: False positives
output:
  #  md_document:
  github_document:
    pandoc_args: --mathjax
#    variant: markdown_github
  preserve_yaml: true
bibliography: [../references/referencesMRbook.bib]
csl: [../references/apa-6th-edition.csl]
---

```{r , echo = FALSE, eval = TRUE}
#knitr::opts_knit$set(base.dir = "/home/rafal/L2/L2projects/backtesting",
knitr::opts_knit$set(base.dir = "/home/rafal/UG/UGprojects/rfl-urbaniak.github.io",
base.url = "https://rfl-urbaniak.github.io/")
knitr::opts_chunk$set(fig.path = "images/")
```




Experts sometimes testify by offering the likelihood ratio as a measure of the strength of the evidence.  An expert, for instance, may testify that the blood-staining on the jacket of the defendant is ten times more likely to be seen if the wearer of the jacket hit the victim (prosecutor's hypothesis) rather than if he did not (defense's hypothesis) [@aitken2010fundamentals, p. 38].  Experts are typically 
advised not to comment  on the posterior odds given the evidence.
As this formulation of the Bayes's theorem makes clear, an assessment of the posterior odds will require a judgment about the prior odds, and the latter lies beyond the competence of an expert. A prominent forensic scientist recommends that  experts `not trespass on the province of the jury by commenting directly on the accused's guilt or innocence, 
\dots and should generally confine their testimony to presenting the likelihood of their evidence under competing propositions' [@aitken2010fundamentals, p. 42]. 

Now, let us follow @aitken2003probability in investigating  its impact on the likelihood ratio of the DNA match. We just add a bit more details to the derivation they present for the sake of clarity. For simplicity we stil assume that the false negative probability is 0, that is, that if the match is real, it will be reported with certainty. We abbreviate:


```{r fig-fpplr,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%", message = FALSE, warning = FALSE}
library(ggplot2)
library(ggthemes)
rmp9 <- 10e-9
rmp3 <- 10e-3
fpp <- seq(0,0.05, by = 0.001)

lr9 <- 1/(rmp9 + (fpp * (1-rmp9)))
lr3 <- 1/(rmp3 + (fpp * (1-rmp3)))

fppTable <- data.frame(fpp,   lr9,  lr3, ref = rep(16.8, length(fpp)))

library(tidyr)
fppTableLong <- gather(fppTable,line,value,c(lr9,lr3,ref), factor_key=TRUE)



ggplot(fppTableLong, aes(x=fpp,y=value, lty = line))+  geom_line()+ylim(c(0,400))+theme_tufte()+ylim(c(0,400))+ylab("Likelihood ratio")+xlab("False positive probability") +scale_linetype_manual(values = c(1,2,3),labels = 
              c(expression(paste("RMP=",10^{-9})),expression(paste("RMP=",10^{-3})),"reference at 16.8"))+ggtitle("Impact of false positive probability on likelihood ratio")+ 
  theme(legend.position = c(0.85,.7))
```




